Reformulating phylogenetic mixed models to improve flexibility and speed

### 2018-11-06 phyloglmm_ms.tex
fixing up discussion

The full single site model using phyloglmm (which matches the single simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except the random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$). 
Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases for \pkg{lme4} but not others.

In general, models that cannot match the simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
PGLM (which lacks the phylogenetic slope parameter) provides reasonable good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation of the intercept ($\beta_0$) estimates slightly undercover the nominal coverage (90\% with 100 species) and have a bad coverage ($<$ 60\%) for the fixed slope parameter.
PGLS, with only one parameter available, confounds all variations (phylogenetic intercept, slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept and overcovering for $\beta_0$, and bad estimates for $\beta_1$.
\mli{refer back to Schielzeth and Forstmeier}

@@

Multiple site fits on the other hand, are much more similar across platforms (as the fitted and simulation models are more similar).
Similar to the single site fits, phyloglmm which matches the simulation model provides good estimates for all parameters except for the random correlation parameter ($\sigma_{sp_{int,slope}}$).
\mli{either keep it or take it out, because  will affect other parameters}
The lack of correlation in \pkg{phyr}'s statistical models does not appear to have a large effect on its ability to estimate other parameters, whereas \pkg{pez} underestimates (lower estimates) many parameters.  
\bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate? Are these singular models? Would they be diagnosable? Can you say anything about when the assumption of independence of slope and intercept \textbf{would} be a problem? (If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
\mli{save it for random slope discussion}

However, despite the similarities of parameter estimates across platforms/methods for the multiple sites simulations fits, the efficiency (measured by computational speed) are different across playforms/methods and sample size.
For example, the median time for \pkg{lme4} to fit 50/100 species model takes $~$ 50/100 seconds versus $~$ 200/2000 seconds for \pkg{pez} and \pkg{phyr}. 
Lastly, it takes $\sim$ 2000 seconds for lme4 to fit 1000 species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 1000 species models where the computational speed did not scale linearly with sample size 25/50/100.
\mli{put a lower bound to give up}

\bmb{Can you put a lower bound on \pkg{pez}'s computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
  or (easier) do a brute-force series of increasing sample size and
  compute a log-log regression of time vs sample size to find the
  approximate power?}

@@

The evolutionary model simulations presented in this paper are an oversimplification of real evolutionary processes in nature, but it is analogous to how researchers fit models to study complicated systems (as it is impossible to create a model to "match" the complexities of nature).
We have fitted models using different platforms/methods that best match the simulation models. 
Models that cannot match the simulation (full) model are expected to underperform; it is important to understand the limitations and performance of these simpler models/assumptions that are restricted by the commonly used methods available.
Using models that include some form of phylogenetic variation and residual variation is necessary to \mli{correctly/robustly?} detect evolutionary process.

@@

Process and observation errors are highly confounded in phylogenetic regression if it is not handled correctly. 
The effects of models that include both process and residual variations (i.e. PGLMM, PGLM) instead of the less realistic model that assumes variation due to evolution and zero observation (PGLS) was large. 
% In particular, models that did not include residual variation tended to over-estimate the amount of variability in the evolutionary process. 
Thus, for single measurement per species model, any method that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e. Pagel's $\lambda$, Blomberg's $K$).
However, if multiple measurements/observations are considered/allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic process by substituting tip variation with residual variation.

\mli{this is specifically targeting boettiger's critque on Pagel's lambda.
https://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html
The difference between the two figures is what it thinks the "tip" variation is. For the second plot, if we treat the sister taxas as "multiple" observation with some observation error on it, then you will get back to the first plot.}

\bmb{again, we/you need to think carefully about what these coarse measures are good for; just avoiding mis-estimating the strength of trait/trait or trait/environment parameters? Is it worth doing some sims on that, or do we basically believe if the total amount of phylogenetic signal is correctly assessed, that we get the right answer for the trait effect? Is \cite{schielzeth2009conclusions} relevant/a useful analogy? (You cite it below, but there are two separate questions: (1) whether including random slopes gives us more insight into evol processes; (2) whether omitting random slopes gives us bad answers even when we're not interested in them.) Worth citing/discussing \cite{uyeda2018rethinking} (as an example of even deeper analysis of evolutionary processes, beyond BM etc.)?}

@@

\bmb{I think you need a lot more here; what does a random slopes model mean? You have one sentence that says ``people should be thinking about random slopes''. Then you talk about correlation (which is a much more subtle/less important issue, I think).}
In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
When analyzing relationships among species between traits and environmental variables, it is entirely plausible that change in effect of predictor variables have on traits are changing in a phylogenetically correlated way.
In fact, it is easier to interpret random slopes in the evolutionary process context (i.e, are size of different species phylogenetically correlated, or growth rate through time of different species phylogenetically correlated?)
\mli{I like to think of it in a longitudinal setting of observing plant length/height. At time zero, they are all still seeds, one can imagine the growth rate are different plant species grow in a phylogenetically correlated way (similar species have similar growth rates).}
It may be easier to simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \cite{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model. 
Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes? and Are we making a mistake by ignoring random slopes \cite{schielzeth2008conclusions}? 
However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.

@@

Our analysis covers the classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models).
Even within the scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, non-BM evolution processes (such as Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes \cite{butler2004phylogenetic}), Bayesian approaches \cite{hadfield2010general}.
More broadly, the simple independence error approach we developed here offers a more efficient and mathematical equivalent way to do Brownian motion evolution process phylogenetic comparative analysis. 
This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as \pkg{MCMCglmm} \cite{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit (i.e. What should we do when we don't have enough data? Should we use a complex model and overfit \cite{barr2013random}, the right balance of complexity and data \cite{baayen2008mixed} or possibility of Bayesian approaches \cite{hadfield2010mcmc}).
More importantly, this implementation in \pkg{lme4} allows users to fit phylogenetic mixed models to the fullest ( that other frequentist platforms lack of (i.e. large data, unbalanced species observations, complex random effects) and explore new ideas.

>>>>

The full single site model using phyloglmm (which matches the single simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except the random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$). 
Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not others.

In general, models that cannot match the simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
PGLM (which lacks the phylogenetic slope parameter) provides reasonable good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation of the intercept ($\beta_0$) estimates slightly undercover the nominal coverage (95\% with 100 species) and have a bad coverage ($<$ 60\%) for the fixed slope parameter.
PGLS, with only one parameter available, confounds all variations (phylogenetic intercept, slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept and overcovering for $\beta_0$, and undercovering for $\beta_1$.

@@

Multiple site fits on the other hand, are much more similar across platforms (as the fitted and simulation models are more similar).
Similar to the single site fits, \pkg{lme4} and \pkg{glmmTMB} which matches the simulation model provides good estimates for all parameters except for the random correlation parameter ($\sigma_{sp_{int,slope}}$).
The lack of correlations in \pkg{pez} and \pkg{phyr} statistical models does not appear to have a large effect on its ability to estimate other parameters, whereas \pkg{pez} underestimates (lower estimates) many parameters.  
% \bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate? Are these singular models? Would they be diagnosable? Can you say anything about when the assumption of independence of slope and intercept \textbf{would} be a problem? (If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
% \mli{save it for random slope discussion}

However, despite the similarities of parameter estimates across platforms for the multiple sites simulation fits, the efficiency (measured by computational speed) are different across platforms and sample size.
For example, the median time for \pkg{lme4} to fit 50/100 species model takes $~$ 50/100 seconds versus $~$ 200/2000 seconds for \pkg{pez} and \pkg{phyr}. 
Lastly, it takes $\sim$ 1000 seconds for lme4 to fit 500 species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500 species for our study, where the computational speed did not scale linearly with sample size 25/50/100.
\pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.

% \bmb{Can you put a lower bound on \pkg{pez}'s computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
%   or (easier) do a brute-force series of increasing sample size and
%   compute a log-log regression of time vs sample size to find the
%   approximate power?}

@@

The evolutionary model simulations presented in this paper are an oversimplification of real evolutionary processes in nature, but it is analogous to how researchers fit models to study complicated systems (as it is impossible to create a model to "match" the complexities of nature).
We have fitted models using different platforms that best match the simulation models. 
Models that cannot match the simulation (full) model are expected to underperform; it is important to understand the limitations and performance of these simpler models that are restricted by the commonly used methods available.
Using models that include some form of phylogenetic variation and residual variation is necessary to detect evolutionary process.
Here are a few points researchers should think about when fitting phylogenetic regression models.

@@

Process and observation errors are highly confounded in phylogenetic regression if it is not handled correctly. 
For single measurement per species model, any method that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e. Pagel's $\lambda$, Blomberg's $K$).
If multiple observations are allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic process by substituting tip variation with residual variation. 
However, one can summarise multiple measurements to a single measurement (for example, mean or median) per species to avoid confounding residual and tip variations.

@@

\bmb{I think you need a lot more here; what does a random slopes model mean? You have one sentence that says ``people should be thinking about random slopes''. Then you talk about correlation (which is a much more subtle/less important issue, I think).}
In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
Neglicting random slopes can lead to erroneous fixed effect estimates \citep{schielzeth2008conclusions} (parameters of interest) as shown in the simulations above. 
When analyzing relationships among species between traits and environmental variables, it is entirely plausible that change in effect of predictor variables have on traits are changing in a phylogenetically correlated way.
In fact, it is easier to interpret random slopes in the evolutionary process context (i.e, are size of different species phylogenetically correlated, or growth rate through time of different species phylogenetically correlated?)
It may be easier to simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model. 
Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes? and Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}? 
% However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.

@@

Our analysis covers the classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models).
Even within the scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, non-BM evolution processes (such as Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes \cite{butler2004phylogenetic}), Bayesian approaches \cite{hadfield2010general}.
More broadly, the simple independence error approach we developed here offers a more efficient and mathematical equivalent way to do Brownian motion evolution process phylogenetic comparative analysis. 
This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as \pkg{MCMCglmm} \citep{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit (i.e. What should we do when we don't have enough data? Should we use a complex model and overfit \citep{barr2013random}, the right balance of complexity and data \citep{baayen2008mixed} or possibility of Bayesian approaches \cite{hadfield2010mcmc}).
More importantly, this implementation in \pkg{lme4} allows users to fit phylogenetic mixed models to the fullest ( that other frequentist platforms lack of (i.e. large data, unbalanced species observations, complex random effects) and explore new ideas.
